Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae

### 2020-11-18 ccsn.tex
Update ccsn.tex
add a PE paragraph in the introduction

\author{Nelson~Christensen}
\affiliation{Artemis, Universit\'e C\^ote d'Azur, Observatoire C\^ote d'Azur, CNRS, CS 34229, F-06304 Nice Cedex 4, France}
\affiliation{Carleton College, Northfield, MN 55057, USA}

@@

\begin{abstract}
The eventual detection of gravitational waves from core-collapse supernovae (CCSN) will help improve our current understanding of the explosion mechanism of massive stars. The stochastic nature of the late post-bounce gravitational wave signal due to the non-linear dynamics of the matter involved and the large number of degrees of freedom of the phenomenon make the source parameter inference problem very challenging. In this paper we take a step towards that goal and present a parameter estimation approach which is based on the gravitational waves associated with oscillations of proto-neutron stars (PNS). Numerical simulations of CCNS have shown that buoyancy-driven g-modes are responsible for a significant fraction of the gravitational wave signal and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations, as demonstrated in~\cite{Torres:2019b}. We use a set of 1D CCSN simulations to build a model that relates the  evolution of the PNS properties with the frequency of the dominant g-mode, which is extracted from the gravitational-wave data using a new algorithm we have developed for our study. The model is used to infer the time evolution of a combination of the mass and the radius of the PNS. The performance of the method is estimated employing simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state. Considering signals embedded in Gaussian gravitational wave detector noise, we show that it is possible to infer PNS properties for a galactic source using Advanced LIGO and Advanced Virgo data at design sensitivities. Third generation detectors such as Einstein Telescope and Cosmic Explorer will allow to test distances of ${\cal O}(100\, {\rm kpc})$. %\tf{Caveats? Gaussian noise?}
% 
%  Core-collapse supernovae are very important phenomena in the Universe and their expected gravitational
%  waves emission is a promising tool to study the onset of the supernova explosion mechanism. The complexity
%  of the gravitational wave signal due to matter effects and the large degrees of freedom of the phenomena
%  makes the source parameter inference problem very challenging. In this paper, we have considered the
%  proto neutron star oscillation modes that carry out a large fraction of the gravitational wave signal and
%  whose frequency time evolution is linked to the system properties through universal relations as
%  demonstrated in ~\cite{Torres:2019b}. We have developed a simple algorithm to extract from the
%  gravitational wave data the frequency evolution of the gravity mode. A set of 1D core collapse
%  supernova simulations is used to build a model of evolution of the proto neutron star properties
%  with the gravity mode frequency. The model is then used to infer the time evolution of a combinaison of
%  the radius and mass of the proto neutron star. We have then estimated the performance of the method by
%  performing simulations with 2D core collapse supernova waveforms covering a progenitor mass range between
%  11 and 40 solar mass and different equations of state. We have shown that for Advanced LIGO and Advanced
%  Virgo detectors at design sensitivity it will be possible to infer proton neutron star properties for
%  a galactic source, while third generation detectors Einstein Telescope and Cosmic Explorer will allow
%  to test distances of several hundreds of kiloparsec.

>>>>

\author{Nelson~Christensen}
\affiliation{Artemis, Universit\'e C\^ote d'Azur, Observatoire C\^ote d'Azur, CNRS, CS 34229, F-06304 Nice Cedex 4, France}
\affiliation{Physics and Astronomy, Carleton College, Northfield, MN 55057, USA}

@@

\begin{abstract}
The eventual detection of gravitational waves from core-collapse supernovae (CCSN) will help improve our current understanding of the explosion mechanism of massive stars. The stochastic nature of the late post-bounce gravitational wave signal due to the non-linear dynamics of the matter involved and the large number of degrees of freedom of the phenomenon make the source parameter inference problem very challenging. In this paper we take a step towards that goal and present a parameter estimation approach which is based on the gravitational waves associated with oscillations of proto-neutron stars (PNS). Numerical simulations of CCSN have shown that buoyancy-driven g-modes are responsible for a significant fraction of the gravitational wave signal and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations, as demonstrated in~\cite{Torres:2019b}. We use a set of 1D CCSN simulations to build a model that relates the  evolution of the PNS properties with the frequency of the dominant g-mode, which is extracted from the gravitational-wave data using a new algorithm we have developed for our study. The model is used to infer the time evolution of a combination of the mass and the radius of the PNS. The performance of the method is estimated employing simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state. Considering signals embedded in Gaussian gravitational wave detector noise, we show that it is possible to infer PNS properties for a galactic source using Advanced LIGO and Advanced Virgo data at design sensitivities. Third generation detectors such as Einstein Telescope and Cosmic Explorer will allow to test distances of ${\cal O}(100\, {\rm kpc})$. %\tf{Caveats? Gaussian noise?}
% 
%  Core-collapse supernovae are very important phenomena in the Universe and their expected gravitational
%  waves emission is a promising tool to study the onset of the supernova explosion mechanism. The complexity
%  of the gravitational wave signal due to matter effects and the large degrees of freedom of the phenomena
%  makes the source parameter inference problem very challenging. In this paper, we have considered the
%  proto neutron star oscillation modes that carry out a large fraction of the gravitational wave signal and
%  whose frequency time evolution is linked to the system properties through universal relations as
%  demonstrated in ~\cite{Torres:2019b}. We have developed a simple algorithm to extract from the
%  gravitational wave data the frequency evolution of the gravity mode. A set of 1D core collapse
%  supernova simulations is used to build a model of evolution of the proto neutron star properties
%  with the gravity mode frequency. The model is then used to infer the time evolution of a combinaison of
%  the radius and mass of the proto neutron star. We have then estimated the performance of the method by
%  performing simulations with 2D core collapse supernova waveforms covering a progenitor mass range between
%  11 and 40 solar mass and different equations of state. We have shown that for Advanced LIGO and Advanced
%  Virgo detectors at design sensitivity it will be possible to infer proton neutron star properties for
%  a galactic source, while third generation detectors Einstein Telescope and Cosmic Explorer will allow
%  to test distances of several hundreds of kiloparsec.

### 2020-11-18 introduction.tex
add a PE paragraph in the introduction

All multidimensional numerical simulations show the systematic appearance in time-frequency diagrams (or spectrograms) of a distinct and relatively narrow feature 
%
%{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow \mab{oscillation} 
%
during the post-bounce evolution of the system, with frequency rising 
from about \unit[100]{Hz} up to a few kHz (at most) and a typical duration of \unit[0.5 -- 1]{s}. This feature has been interpreted as a continuously excited gravity mode (g-mode, see \citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS \citep{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015, Kuroda:2016, Andresen:2017}. 
In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS. The {typical} frequencies of {these} modes make them interesting targets for
%promising {source} for 
ground-based GW interferometers.

@@

>>>>

All multidimensional numerical simulations show the systematic appearance in time-frequency diagrams (or spectrograms) of a distinct and relatively narrow feature 
%
%{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow \mab{oscillation} 
%
during the post-bounce evolution of the system, with frequency rising 
from about \unit[100]{Hz} up to a few kHz (at most) and a typical duration of \unit[0.5 -- 1]{s}. This feature has been interpreted as a continuously excited gravity mode (g-mode, see \citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS \citep{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015wra, Kuroda:2016, Andresen:2017}. 
In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS. The {typical} frequencies of {these} modes make them interesting targets for
%promising {source} for 
ground-based GW interferometers.

@@

Many previous data analysis efforts focused on the reconstruction of the gravitational wave strain amplitude
without assuming a particular signal model \citep{Summerscales:2008,Klimenko:2015ypf,CornLitt}. As numerical
simulations became more available other methods using catalogs of waveforms have been proposed to
identify the supernova explosion mechanism.
Among them, principal component analysis helps at reducing the complexity of CCSN waveform catalogs to fewer
parameters \citep{Heng:2009,roever:09,Edwards:2014,powell:2016,powell:2017,Roma:2019kcd}.
The classification challenge is also well addressed with deep learning methods \citep{astone:2018,Chan:2019fuz}.
